\batcave: Calling somatic mutations with a tumor- and site-specific prior

### 2020-01-10 ResponseToReviews.tex

When we reanalyzed our test data using this new approach, we did detect a profile shift in the AML data.
Nevertheless, BATCAVE still performed extremely well.
\end{response}

Related to this, there is probably a minimum number of high-confidence mutations that are required by BATCAVE to produce reliable results. As with the previous point, ideally the batcaver should address or warn the user if the assumptions are violated.
\begin{response}
Our simulation studies suggest that a few hundred high-confidence mutations are enough to reliably infer even a sharply peaked mutation profile (Fig. 3).
We have modified \texttt{batcaver} so that it now emits a warning if there are fewer than 500 high-confidence mutations in a sample.
\end{response}

Lastly, as correctly pointed out by the second reviewer, the sequencing coverage only makes sense in the context of the purity of the samples and this information should be provided to the reader.
\begin{response}
The direct effect of contamination by normal cells on the BATCAVE algorithm is to bias the estimated allele frequencies that are inputs to Eq.~11 for estimating the overall mutation rate.
This bias can be simply corrected given an estimate of tumor purity.
We have thus added a \texttt{--tumor\_purity} option to \texttt{batcaver}.
When this is enabled, the estimated allele frequencies will be appropriately adjusted before applying Eq.~11.
See new text in Methods for details.

We also reanalyzed our real data examples, using purity estimates from the original publications.
Accounting for tumor purity changed the quantitative results only slightly, and did not change the qualitative conclusions.
We have updated the text, Table~1, and Fig.~4E\&F to reflect these new results.

### 2020-01-10 NAR_DRAFT.tex
add comparison figure
minor change to the main table. both aml and cell got slightly worse
fix supplementary figure width
Changes to text for updated results. Also some to Response.

\begin{table*}[b]
  \tableparts{%
  \caption{Variant calling metrics for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular}{lllllllllll}
  \toprule
  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .995 & .996 & -- & --
  \\
  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .972 & .972 & -- & --
  \\
  \botrule
  \end{tabular}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
 \end{table*}

@@

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $4\cdot10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
\rngcomment{
For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
The low-confidence mutation profile contains additional spikes of mutations corresponding to XXX (Fig.~\ref{fig:AML_profiles}).
}
Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
\rngcomment{Need to update!}
We estimated an average per-base mutation rate for these tumor regions of $4\cdot10^{-8}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

@@

\begin{figure} [H]
\centering
  %\includegraphics{XXX}
  \caption{Mutation profiles of high-confidence and low-confidence mutations in the AML data.}
\label{fig:AML_profiles}
\end{figure}

>>>>

\begin{table*}[b]
  \tableparts{%
  \caption{Variant calling metrics for all data sets.}
  \label{table:01}%
  }{%
  \begin{tabular}{lllllllllll}
  \toprule
  Scenario & Mutation profile & $\mu$ & \multicolumn{2}{c}{AUROC} & \multicolumn{2}{c}{AUPRC} & \multicolumn{2}{c}{ICI} 
  \\
  & & (estimated) & MuTect & \batcave & MuTect & \batcave & MuTect & \batcave
  \\
  \colrule
  100X whole genome & Concentrated & 3.6e-7 & .987 & .993 & .972 & .975 & .117 & .287
  \\
  100X whole genome & Intermediate & 3.2e-7 & .987 & .989 & .972 & .973 & .118 & .214
  \\
  100X whole genome & Diffuse & 3.2e-7 & .988 & .989 & .971 & .973 & .120 & .219
  \\
  500X whole exome & Concentrated & 3.6e-7 & .848 & .929 & .674 & .758 & .138 & .109
  \\
  500X whole exome & Intermediate & 3.6e-7 & .847 & .881 & .677 & .706 & .108 & .112 
  \\
  500X whole exome & Diffuse & 3.6e-7 & .850 & .873 & .676 & .698 & .105 & .116
  \\
  real AML \cite{Griffith2015} & Actual & 3.6e-8 & -- & -- & .988 & .996 & -- & --
  \\
  real breast \cite{Shi2018} & Actual & 6.6e-7 & -- & -- & .968 & .972 & -- & --
  \\
  \botrule
  \end{tabular}%
  }
  {$\mu$ = per-base mutation rate, AUROC/AUPRC = Area Under Receiver Operating Characteristic / Precision-Recall Curve, ICI = Integrated Calibration Index. Smaller values of ICI are superior.}
 \end{table*}

@@

Griffith et al.\ sequenced the whole genome of an acute myeloid leukemia (AML) primary tumor to a depth of $>$360X and used targeted sequencing to validate nearly 200,000 mutations~\cite{Griffith2015}.
We estimated a per-base mutation rate for this tumor of $3.6\times10^{-8}$, which is consistent with previous estimates of AML mutation rates \cite{Griffith2015,Williams2018}.
\rngcomment{
For this tumor, the mutation profiles of high- and low-confidence mutations differ ($p \sim 10^{-12}$).
The low-confidence mutation profile contains additional spikes of mutations corresponding to five substitution types (Fig.~\ref{fig:AML_profiles}).
}
Nevertheless, for both MuTect and \batcave, the precision-recall curve is almost perfect for the validated variants (.995 \& .996 area under the curve) (Fig.~\ref{NAR-wes_fig}E and Table~\ref{table:01}).

Shi et al.\ performed multi-region whole exome sequencing on six individual breast tumors to a mean target sequencing depth of 160X and validated all variants identified by three different variant calling pipelines~\cite{Shi2018}.
%We downloaded aligned reads for this tumor, and called variants with MuTect 1.1.7 as described in Materials \& Methods.
\rngcomment{Need to update!}
We estimated an average per-base mutation rate for these tumor regions of $6.6\times10^{-7}$, which is consistent with observed mutation rates for breast cancers \cite{Alexandrov2019} and with the low number of validated somatic mutations.
For the validated variants, MuTect and \batcave yielded almost identical precision-recall curves (Fig.~\ref{NAR-wes_fig}F and Table~\ref{table:01})

@@

\begin{figure} [H]
\centering
  \includegraphics[width=\textwidth]{figures/aml_sig_comp.pdf}
  \caption{Mutation profiles of A) high-confidence and B) low-confidence mutations in the AML data.}
\label{fig:AML_profiles}
\end{figure}

\clearpage
